---
title: "Taxa Plots"
execute:
  echo: false
  message: false
  warning: false
---

```{r, include=FALSE}
# Load functions in "common_code.R"
# at some point this should be moved into bookdown
source("src/common_code.R", local = knitr::knit_global())
```

```{r}
physeq.prev <- readRDS('results/phyloseq.prevfiltered.notglommed.RDS')
```


## Taxa plots for Coliforms

Based on my search, coliform bacteria belong to Enterobacteriaceae family and include species of the following genera: Citrobacter, Enterobacter, Escherichia, Hafnia, Klebsiella, Serratia, and Yersinia, so I filtered the data to only include these genera and then created violin plots for those. If you can give me a list of bacteria at any level you are interested in, I can create a plot for it.


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=11, fig.height=7}

set.seed(1)
n <- 230
palette <- distinctColorPalette(n)

Level <- "Genus"

# Calculate relab for samples -----
#(tax.level <- subset_taxa(physeq.prev, Family=="Enterobacteriaceae"))

tax.level <- subset_taxa(physeq.prev, Genus %in% c("Citrobacter", "Enterobacter", "Escherichia", "Hafnia", "Klebsiella", "Serratia", "Yersinia"))
tax.level <- tax_glom(tax.level, Level, NArm = F)
tax.level.rel <- microbiome::transform(tax.level, "compositional")
# Agglomerate at taxa level -----
tax.level.rel <- aggregate_rare(tax.level.rel, level = Level, detection = 0.1/100, prevalence = 5/100)

## rename samples
taxa_names(tax.level.rel) <- make.unique(as.vector(as.data.frame(tax_table(tax.level.rel))[[Level]]))

# Grab taxa info
data <- t(as.data.frame(otu_table(tax.level.rel)))
data <- rownames_to_column(as.data.frame(data), "Sample_ID")

# Grab metadata info
mapping_frame <- sample_data(tax.level.rel)
mapping_frame <- as.data.frame(as.matrix(mapping_frame))

# Combine and prep for plot
final <- merge(data, mapping_frame, by="Sample_ID")
final_tab <- reshape2::melt(final)

# Make a quick section for boxplot plotting
  p <- ggplot(data = final_tab, aes(x = fct_reorder(variable, value, .fun = median, .desc =TRUE), y=value, fill = variable)) 
  
  p$data$group <- factor(p$data$group,levels = c(
                  "Drags_PBS",
                   "Drags_BPW",
                   "Booties_PBS",
                   "Booties_BPW", 
                   "Grabs_N/A"))
  p + 
    #geom_boxplot() + 
      geom_violin() + 
    scale_fill_brewer(palette = "Set2", drop=F) +
    theme_bw() + 
    #ggtitle(paste(microbe, Level, "Box Plot"), subtitle = sublabl) 
    #facet_grid(get(group)~.) + 
    xlab("Coliforms")+
    ylab("Relative Abundance")+
     guides(fill = guide_legend(title = "Genus"))+
    facet_wrap(.~group,nrow = 1)+
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size=12), plot.margin = unit(c(0,0,0,1.25), "cm"),
          plot.subtitle=element_text(size=9, hjust=0.5, face="italic", color="black"),
          legend.text=element_text(size=12),
          strip.text.x = element_text(size = 14),
          legend.title = element_text(size = 14, face = "bold"),
          axis.title = element_text(size = 14, face = "bold"))

```

